---
title: "ADAE"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(sdtm.terminology)
```

::: {.callout-note}
### Session at a Glance

**Total core time: about 85 minutes.**

| Section | Time | Type |
|---|---|---|
| When Do You Use This? + Learning Objectives | 5 min | Reading |
| Background: counting events, not measuring them | 20 min | Reading |
| Example 1: The ADAE spine and treatment emergence | 25 min | Worked example |
| Example 2: Occurrence flags and the denominator | 20 min | Worked example |
| What Can Go Wrong | 10 min | Reading |
| Exercises + Comprehension Check | 5 min | Practice / Self-test |

The course's first OCCDS dataset, and the first session that reads an
ADaM dataset as an input. For self-paced study, split after Example 1.
:::

::: {.callout-tip}
### How to Use the Code in This Session

Every code chunk below is written so you can copy it into your R
console (or an R script) and run it yourself, in the order shown. Run R
from the project root, as in earlier sessions.

This session reads `data/sdtm/ae.csv` and — for the first time —
`data/adam/adsl.csv`. Treatment emergence cannot be decided without the
treatment dates ADSL carries, which is why ADSL came first.
:::

## When Do You Use This? {#sec-adae}

::: {.callout-tip}
A safety table needs every adverse event classified as treatment
emergent or not, because an event that began before the first dose is
not evidence about the drug. A reviewer asks which subjects had their
first *severe* event while on treatment. Someone needs the percentage of
subjects who experienced nausea — and gets a different answer depending
on what they divide by.

None of those questions is about a measured *value*. There is no result
to record, no unit, no change from baseline. There is only a list of
things that happened, and questions about which of them count. That
raises the question this session answers: **what does an analysis
dataset look like when there is nothing to measure — only occurrences to
count?**

Builds on [ADaM Concepts](adam-concepts.qmd), [ADSL](adsl.qmd), and the
SDTM AE domain from [Events](sdtm-events.qmd).
:::

## Learning Objectives

After completing this session you will be able to:

- Recognise when OCCDS rather than BDS is the right analysis structure,
  and say why in terms of what is being analysed
- Build an ADAE from SDTM AE with the subject-level variables merged
  from ADSL
- Derive `ASTDT`, `ASTDY` and `TRTEMFL`, and explain why treatment
  emergence is an ADaM derivation rather than an SDTM fact
- Derive occurrence flags, and identify which ones dictionary licensing
  puts out of reach
- Take an incidence denominator from ADSL rather than from the event
  data, and explain what goes wrong otherwise

## Background: Counting Events, Not Measuring Them

*Estimated time: ~20 minutes (reading)*

::: {.callout-important}
### The one thing to remember

**ADAE is one row per adverse event, carried from SDTM and answered
against the treatment window: not what value was measured, but whether
this event counts for this analysis — was it treatment emergent, and is
it this subject's first of its kind?**
:::

Every ADaM dataset you have met so far has been built around a measured
value. In BDS there is `AVAL`, a `PARAM` that says what it is, and a
baseline to compare it against. Adverse events have none of that. You do
not measure an adverse event; you record that one occurred. Analysis
then *counts* occurrences.

That difference is not cosmetic, and CDISC gives it its own structure
and its own document. The **Occurrence Data Structure (OCCDS)** applies
when three things hold (OCCDS v1.1, §1.1.2):

1. There is **no need for `AVAL` or `AVALC`**, because "occurrences are
   counted in analysis."
2. A **dictionary hierarchy** is typically used for coding, and
   "remapping this hierarchy to BDS variables PARAM and generic \*CAT
   variables would lose the structure and meaning of the dictionary."
3. The data content is **typically not modified** for analysis.

The guide summarises the division in one sentence: "typically, findings
data fit nicely into BDS, and events and interventions fit nicely into
OCCDS" (OCCDS v1.1, §1.1.2). And it states the structural consequence
bluntly — in OCCDS "there is no `PARAM` nor `AVAL`, for example" (OCCDS
v1.1, §3.2). If you go looking for the variables that organised the last
session, they are not here, and their absence is the design.

::: {.callout-note}
### A note on versions worth understanding

OCCDS v1.1's own front matter says it is "based on the Analysis Data
Model (ADaM) v2.1 and the ADaM Implementation Guide (ADaMIG) v1.2," and
it cites SDTMIG v3.3 section numbers throughout. This course targets
ADaMIG v1.3 and SDTMIG v3.4, so that looks like a mismatch.

It is not. **OCCDS v1.1 is CDISC's designated companion to ADaMIG
v1.3** — the ADaMIG v1.3 documentation points to OCCDS v1.1 for
occurrence data, and the v1.3 release package ships it. CDISC's
foundational standards are versioned on staggered cycles: a companion
document is not reissued merely because the document it references moved
on. Read the front matter as telling you when OCCDS was last revised,
not as telling you which ADaMIG to pair it with.

This is worth internalising because you will meet it again. "The version
numbers do not match" is rarely the same thing as "these documents do
not go together."
:::

### The adverse event SubClass, and an honest constraint

OCCDS defines a **SubClass ADVERSE EVENT** for the standard AE analyses.
A dataset in it must have Class `OCCURRENCE DATA STRUCTURE`, is usually
one record per record in the corresponding SDTM domain, and always takes
AE as its SDTM input, with additional information from SUPPAE, FA and
ADSL (OCCDS v1.1, §3.1.2).

It also carries a requirement this course cannot meet:

> For the SubClass ADVERSE EVENT, it is required that all levels of
> terms for the primary path in the MedDRA hierarchy — System Organ
> Class (SOC), High Level Group Term (HLGT), High Level Term (HLT),
> Lowest Level Term (LLT), and Preferred Term (PT) — be included. For
> other OCCDS datasets, these variables are recommended… but not
> required (OCCDS v1.1, §3.2.3).

GLPX-001 has no MedDRA coding at all. `AEDECOD` is null on every row,
because MedDRA is a licensed dictionary and this course does not
fabricate coded terms — the same discipline that keeps it from inventing
lab reference ranges or a half-life for a fictional compound. So the
SubClass is out of reach.

The right response is not to fake the codes, and not to abandon the
dataset. It is to build ADAE with Class `OCCURRENCE DATA STRUCTURE` and
no SubClass — and the guide is unusually direct that this is correct
rather than merely tolerated:

> Data for a particular study that could have been coded but was not
> **should use this structure**, because analysis results are similar,
> and this will allow analysis programming to work the same way (OCCDS
> v1.1, §1.1.2).

That sentence is doing real work. It says uncoded occurrence data
belongs in OCCDS, so that the analysis code you write today keeps
working when the data *is* coded tomorrow. The supporting point is in
§3.1.2: "not all OCCDS datasets that contain adverse event data will
necessarily be of SubClass ADVERSE EVENT." Uncoded is a legitimate
state, not a broken one.

What it costs is specific, and it is worth working out precisely rather
than by feel. The occurrence flags sit in two tables — the general ones
in Table 3.2.6.1, and a second table titled *MedDRA Occurrence Flag
Variables* (OCCDS v1.1, §3.2, Table 3.2.6.2). Whether we can build each
one depends on what its derivation *keys on*:

| Flag | Keys on | Where CDISC files it | Buildable here |
|---|---|---|---|
| `AOCCFL` | first treatment-emergent record | Table 3.2.6.1 | **Yes** |
| `AOCCIFL` | maximum severity (`ASEVN`) | Table 3.2.6.1 | **Yes** |
| `AOCCPFL` | preferred term (`--DECOD`) | Table 3.2.6.1 | No |
| `AOCCPIFL` | preferred term + severity | Table 3.2.6.1 | No |
| `AOCCSFL` | system organ class | Table 3.2.6.2 | No |
| `AOCCSIFL` | system organ class + severity | Table 3.2.6.2 | No |

Read the middle column and the boundary is obvious. Read the right-hand
one and it is not — **the line falls across the two tables, not along
them.** `AOCCPFL` and `AOCCPIFL` sit in the *general* table, yet we
cannot build them, because their derivations key on `--DECOD`. The
table titled "MedDRA" contains some of what a dictionary costs you, not
all of it.

So the rule to carry away is about derivations, not headings: **the
buildable set is bounded by which coded terms we lack — preferred term
and system organ class — and that boundary cuts across CDISC's table
structure rather than following it.** `AOCCFL` and `AOCCIFL` survive
because they key on record order and on `ASEVN`, which come from
severity and dates that GLPX-001 has. Everything keyed on a dictionary
term goes, wherever it happens to be filed.

That is the general habit worth building. When a constraint bites —
licensing, a domain you did not collect, a variable the vendor never
sent — work out what each derivation *depends on*. Grouping in the
standard is a useful signpost and an unreliable rule.

### Treatment emergence: the question ADAE exists to answer

An adverse event that began before the first dose tells you about the
subject, not the drug. **`TRTEMFL`** is the flag that makes the
distinction, and it is the central derivation of this dataset. The IG
gives the derivation directly (OCCDS v1.1, §3.2, Table 3.2.5.3):

> If `ADSL.TRTSDT <= ASTDT <= ADSL.TRTEDT + x days` then
> `TRTEMFL = "Y"`

where "the number *x* in this derivation is defined by the producer and
often incorporates the known half-life of the drug." `TRTEMFL` is
Conditional in OCCDS generally and Required for the SubClass; for a
dataset outside the SubClass it is "conditional on whether the concept
of treatment emergent is a key feature of the analysis," which for a
safety dataset it plainly is.

Notice what that derivation depends on: three dates. **No coded term
appears anywhere in it.** Treatment emergence is a pure timing question,
which is why it remains fully available to us when the dictionary-based
flags are not.

Notice also where two of those three dates come from: `ADSL`. You cannot
decide whether an event was treatment emergent from the AE domain alone
— SDTM AE does not know when the subject was dosed. This is the first
time in the course that an ADaM dataset is an *input* to another ADaM
dataset, and it is not incidental. It is why ADSL is built first and why
it is required in a submission even when nothing else is.

::: {.callout-note}
### `TRTEMFL` is not `AETRTEM`

SDTM can carry its own treatment-emergent flag as a Supplemental
Qualifier, `AETRTEM` in SUPPAE (OCCDS v1.1, §3.2, Table 3.2.5.3). Where
it exists, including it alongside `TRTEMFL` adds traceability — but the
guide is explicit that "`TRTEMFL` may differ from `AETRTEM` due to
different definitions, date imputation, and other analysis rules."

This is the same lesson as ADSL's population flags, which ADaM
re-derives rather than copying from SDTM's Supplemental Qualifiers
(ADaMIG v1.3, §3.5). The analysis flag is the analysis's own. GLPX-001
has no SUPPAE, so `AETRTEM` simply does not exist here — stated rather
than invented.
:::

### Timing, and what a missing date is allowed to mean

`ASTDT` ("Analysis Start Date") is "created from converting `XX.--STDTC`
from character ISO 8601 format to numeric date format, applying
imputation rules if specified in the statistical analysis plan (SAP) or
metadata"; `AENDT` follows from `--ENDTC` (OCCDS v1.1, §3.2, Table
3.2.4.1). The original `--STDTC` and `--ENDTC` are carried across
unchanged alongside them, for traceability.

Two consequences follow, and both matter.

First, **imputation is a documented decision, not a default.** If a
partial or missing date is imputed, `ASTDTF` records the level of
imputation, from the `DATEFL` codelist, and it is conditional precisely
"on whether `ASTDT` (or the date part of `ASTDTM`) was imputed" (OCCDS
v1.1, §3.2, Table 3.2.4.1). No imputation rule, no imputation — and
`ASTDTF` stays null because nothing was imputed.

Second, `ASTDY` ("Analysis Start Relative Day") uses a rule you have
already met:

> `ASTDT - ADSL.TRTSDT + 1` if `ASTDT >= ADSL.TRTSDT`, else
> `ASTDT - ADSL.TRTSDT` (OCCDS v1.1, §3.2, Table 3.2.4.1)

That is the no-day-zero convention from SDTMIG v3.4 §4.4.4 — the same
two-branch shape you implemented in [The DM Domain](sdtm-dm.qmd). But
the anchor has moved, and OCCDS says so in the variable's own
definition: `ASTDY` is "the number of days from an anchor date **(not
necessarily `DM.RFSTDTC`)** to `ASTDT`" (OCCDS v1.1, §3.2, Table
3.2.4.1).

SDTM's `--DY` counts from `DM.RFSTDTC`; ADaM's `ASTDY` counts from
whatever the analysis nominates, here `ADSL.TRTSDT`. The guide adds that
`ASTDY` "may differ from `--STDY` due to date imputation and the option
in ADaM to use a reference date other than SDTM's `RFSTDTC`," and
recommends carrying `AESTDY` alongside when it exists, so both are
visible. Same arithmetic, deliberately different origin, because
analysis gets to choose its own reference point — and must then say
which one it chose.

That parenthesis closes an arc this course has been building since
[The DM Domain](sdtm-dm.qmd). The study-day rule is fixed; the day it
counts from is not.

### Where denominators come from

One sentence in OCCDS is worth more than any variable definition in it
(OCCDS v1.1, §3.2.1):

> Be aware that **only subjects with a record in an input SDTM domain
> would have a record in the analysis dataset**. For this reason, it is
> recommended that population indicators and denominator counts for
> percentages be derived from ADSL and not from the occurrence analysis
> dataset.

A subject who had no adverse events has no row in ADAE. They still
belong in the denominator of every incidence percentage, because they
were dosed and observed and nothing happened. Compute "percent of
subjects with nausea" from the event dataset and you have silently
restricted the population to subjects who had *some* event — inflating
every percentage in the table. Example 2 puts real numbers on how large
that error is.

## Example 1: The ADAE Spine and Treatment Emergence

*Estimated time: ~25 minutes (worked example)*

Start with both inputs. This is the shape of every OCCDS build: the
event records from SDTM, and the subject-level context from ADSL.

```{r read-inputs}
#| eval: true

ae <- read_csv("data/sdtm/ae.csv", show_col_types = FALSE,
               col_types = cols(.default = col_character()))

adsl <- read_csv("data/adam/adsl.csv", show_col_types = FALSE)

c(ae_rows          = nrow(ae),
  subjects_with_ae = n_distinct(ae$USUBJID),
  adsl_subjects    = nrow(adsl))
```

Hold on to those three numbers. 389 event records belong to 262
subjects, but the trial has 400. The 138 subjects who had no adverse
event are absent from AE entirely — and that gap is the denominator
problem, visible before we have derived anything.

Now merge the subject-level variables and derive the analysis dates.
`AESEQ` is what makes each row traceable back to its source: where rows
come from a single SDTM domain, `--SEQ` supplies the whole row
identifier, and the `SRCDOM`/`SRCSEQ` pair is the alternative for
multi-domain sources (OCCDS v1.1, §3.2.2).

```{r adae-spine}
#| eval: true

adae_spine <- ae |>
  left_join(
    adsl |> select(USUBJID, SUBJID, SITEID, TRT01A, TRTSDT, TRTEDT, SAFFL),
    by = "USUBJID"
  ) |>
  mutate(
    AESEQ = as.integer(AESEQ),

    ASTDT = as.Date(AESTDTC),          # OCCDS §3.2, Table 3.2.4.1
    AENDT = as.Date(AEENDTC),
    ASTDTF = NA_character_,            # nothing imputed, so nothing to flag

    ASTDY = if_else(ASTDT >= TRTSDT,   # no day 0; anchored to ADSL.TRTSDT
                    as.integer(ASTDT - TRTSDT) + 1L,
                    as.integer(ASTDT - TRTSDT))
  )

adae_spine |>
  select(USUBJID, AESEQ, AETERM, AESTDTC, ASTDT, TRTSDT, ASTDY) |>
  head(4)
```

Note that `AEDECOD` came across in the join and is null throughout —
that is the licensing constraint, present in the data rather than
described in the abstract:

```{r decod-null}
#| eval: true

c(rows = nrow(adae_spine),
  aedecod_populated = sum(!is.na(adae_spine$AEDECOD)),
  distinct_aeterm   = n_distinct(adae_spine$AETERM))
```

25 distinct verbatim terms, zero coded terms. Every analysis that needs
a preferred term or a system organ class is unavailable, and every
analysis that needs only timing and severity is unaffected.

### Severity: the same-name rule, once more

`AESEV` is copied from SDTM untouched. The analysis version — recased
for presentation — goes in `ASEV`, with `ASEVN` as its numeric
counterpart, and the guide notes low intensity should correspond to a
low value (OCCDS v1.1, §3.2, Table 3.2.8.1).

```{r severity}
#| eval: true

adae_spine <- adae_spine |>
  mutate(
    ASEV  = recode(AESEV, MILD = "Mild", MODERATE = "Moderate",
                   SEVERE = "Severe"),
    ASEVN = case_when(ASEV == "Mild"     ~ 1L,
                      ASEV == "Moderate" ~ 2L,
                      ASEV == "Severe"   ~ 3L)
  )

count(adae_spine, AESEV, ASEV, ASEVN)
```

This is the "same name, same meaning, same values" rule from
[ADSL](adsl.qmd) doing its job (ADaMIG v1.3, §3.1.1, item 3). We wanted
mixed case for the tables; `AESEV` carries an SDTM name, so it may not
be touched, and the recased values go somewhere else. The controlled
terminology confirms why that matters:

```{r ct-check}
#| eval: true

sev_code <- (ct("list") |> filter(term == "AESEV"))$clst_code[1]
ny_code  <- (ct("list") |> filter(term == "NY"))$clst_code[1]

tibble(
  variable = c("AESEV", "ASEV", "AESER"),
  codelist = c(sev_code, sev_code, ny_code),
  values   = c("MILD/MODERATE/SEVERE", "Mild/Moderate/Severe", "N/Y"),
  all_valid = c(
    all(is_term(unique(adae_spine$AESEV), clst_code = sev_code)),
    all(is_term(unique(adae_spine$ASEV),  clst_code = sev_code)),
    all(is_term(unique(adae_spine$AESER), clst_code = ny_code))
  )
)
```

::: {.callout-tip}
### Run It Yourself

`AESEV` and `AESER` come back `TRUE`. **`ASEV` comes back `FALSE`, and
that is the correct result.** OCCDS marks `ASEV` with an asterisk
meaning "variable may be subject to producer-defined controlled
terminology" (OCCDS v1.1, §3.2, Table 3.2.8.1, footnote b). `Mild` is
not a CDISC submission value and was never meant to be — it is the
sponsor's presentation choice, defined in the metadata rather than
checked against a codelist.

This is the same point `AGEGR1` made in [ADSL](adsl.qmd) and `PARAMCD`
made in [ADaM Concepts](adam-concepts.qmd). A `FALSE` from `is_term()`
means "not CDISC CT," which is a finding only if the variable was
supposed to be CDISC CT. Knowing which variables those are is the skill.
:::

### The flag itself

Now treatment emergence. We must choose the *x* in `TRTEDT + x days`:

```{r trtemfl}
#| eval: true

TRTEM_WINDOW_DAYS <- 0L   # sponsor decision — see the note below

adae_spine <- adae_spine |>
  mutate(
    TRTEMFL = case_when(
      is.na(ASTDT) ~ NA_character_,
      ASTDT >= TRTSDT & ASTDT <= TRTEDT + TRTEM_WINDOW_DAYS ~ "Y",
      .default = "N"
    )
  )

count(adae_spine, TRTEMFL)
```

::: {.callout-important}
### Two honest results in one table

**On *x* = 0.** The IG says *x* "often incorporates the known half-life
of the drug" — a real trial commonly counts events for some days after
the last dose, because the drug is still present. GLPX-001 is a
fictional compound, and inventing a half-life for it would be inventing
a pharmacological fact to dress up an example. So the window here is
exactly first dose to last dose inclusive, declared as a sponsor
decision. **Do not read *x* = 0 as the norm.** In a real study *x* is a
protocol- or SAP-level decision with a clinical justification behind it,
and it is often greater than zero.

**On the absence of any `"N"`.** Every one of the 388 datable events in
GLPX-001 falls inside the treatment window — the earliest begins one day
after first dose, the latest a week before last dose. So the flag comes
out `"Y"` for all of them, and this trial contains **no non-emergent
event to show you**. The simulator generated on-treatment events only.

That is a limitation of the data, not of the concept, and the honest
move is to say so rather than manufacture a pre-treatment event to make
a prettier demonstration. Real trials routinely have them: an event
recorded at screening, a condition present before the first dose. The
derivation above would assign those `"N"`, and the safety tables would
exclude them. In GLPX-001, *x* = 0 and *x* = 30 would give identical
results — which itself tells you the window choice cannot be validated
against this dataset, only declared.
:::

### The event with no start date

One row behaves differently, and it is the defect you met in
[Events](sdtm-events.qmd):

```{r missing-date}
#| eval: true

adae_spine |>
  filter(is.na(ASTDT)) |>
  select(USUBJID, AESEQ, AETERM, AESTDTC, ASTDT, ASTDY, ASTDTF, TRTEMFL)
```

::: {.callout-tip}
### Run It Yourself

Subject `GLPX001-103-199` has an adverse event — constipation — with no
recorded start date. Follow the consequences across the row:

- `ASTDT` is null, because there was nothing to convert and **no
  imputation rule was specified** for this study.
- `ASTDTF` is null too, and that is the correct value. The imputation
  flag records *the level of imputation applied*. Nothing was imputed,
  so there is nothing to record. A populated `ASTDTF` here would be a
  lie about what we did.
- `ASTDY` is null, because it is arithmetic on a date we do not have.
- `TRTEMFL` is null — **not `"N"`**.

That last one is the judgement call, and it is worth being deliberate
about. `"N"` asserts that the event was *not* treatment emergent. We do
not know that. Absence of a date is not evidence of a date outside the
window; it is absence of evidence. Coding it `"N"` would quietly move an
unknown into the "not the drug's problem" column, which is exactly the
direction of error a safety analysis must not make by accident.

Null says what is true: this event cannot be classified. A real study
would resolve it — query the site for the date, or apply the SAP's
imputation rule and set `ASTDTF` accordingly. Both are decisions someone
must make and document. Neither is a default.
:::

## Example 2: Occurrence Flags and the Denominator

*Estimated time: ~20 minutes (worked example)*

Occurrence flags answer "is this row the first of its kind for this
subject?" They exist because a subject who reports nausea four times
should count once in an incidence table. The derivation given for
`AOCCFL` is "sort the data in the required order and flag the first
treatment emergent record for each subject" (OCCDS v1.1, §3.2, Table
3.2.6.1) — so the sort comes first, and the flags key on `TRTEMFL`.

```{r occurrence-flags}
#| eval: true

adae <- adae_spine |>
  arrange(USUBJID, ASTDT, AESEQ) |>
  group_by(USUBJID) |>
  mutate(
    te_     = !is.na(TRTEMFL) & TRTEMFL == "Y",
    maxsev_ = if (any(te_)) max(ASEVN[te_], na.rm = TRUE) else NA_integer_,
    ismax_  = te_ & !is.na(ASEVN) & !is.na(maxsev_) & ASEVN == maxsev_,

    AOCCFL  = if_else(te_ & cumsum(te_) == 1L, "Y", NA_character_),
    AOCCIFL = if_else(ismax_ & cumsum(ismax_) == 1L, "Y", NA_character_)
  ) |>
  ungroup() |>
  select(-te_, -maxsev_, -ismax_)

c(AOCCFL_Y  = sum(adae$AOCCFL  == "Y", na.rm = TRUE),
  AOCCIFL_Y = sum(adae$AOCCIFL == "Y", na.rm = TRUE),
  subjects_with_te_event = n_distinct(adae$USUBJID[!is.na(adae$TRTEMFL)]))
```

261 subjects have at least one treatment-emergent event, and each gets
exactly one `AOCCFL` and one `AOCCIFL`. Not 262 — subject `103-199`,
whose only event has no date, has no classifiable event and therefore no
first occurrence to flag. The missing date propagates all the way to the
occurrence flags, which is what a missing date should do.

The two flags are easiest to understand on a subject with several
events:

```{r one-subject}
#| eval: true

adae |>
  filter(USUBJID == "GLPX001-101-147") |>
  select(AESEQ, AETERM, ASTDT, ASTDY, ASEV, ASEVN, TRTEMFL, AOCCFL, AOCCIFL)
```

::: {.callout-note}
### Reading the Output Line by Line

| What you see | What it means |
|---|---|
| Four events, all `TRTEMFL = "Y"` | All began during the treatment window. |
| `AOCCFL = "Y"` on the diarrhoea at day 51 | The subject's **first** treatment-emergent event, by date. This is the row an "any adverse event" incidence count uses. |
| `AOCCIFL = "Y"` on the nasopharyngitis at day 134 | The subject's first event at their **maximum severity** — the only Moderate one among four. This is the row a "worst severity per subject" table uses. |
| The two flags on **different rows** | The point of having both. "First event" and "worst event" are different questions, and a table that answers one must not be built from the other's flag. |

Had this subject's severities run the other way — Moderate first — both
flags would have landed on the same row. That they can coincide is
exactly why the distinction is easy to miss and worth deriving
explicitly.
:::

### The denominator

Now the error OCCDS warns about (OCCDS v1.1, §3.2.1). Here are the same
subjects counted two ways:

```{r denominator}
#| eval: true

with_ae <- adae |>
  filter(TRTEMFL == "Y") |>
  distinct(USUBJID, TRT01A) |>
  count(TRT01A, name = "subjects_with_ae")

safety_n <- adsl |>
  filter(SAFFL == "Y") |>
  count(TRT01A, name = "safety_population")

with_ae |>
  left_join(safety_n, by = "TRT01A") |>
  mutate(
    correct = round(100 * subjects_with_ae / safety_population, 1),
    wrong   = round(100 * subjects_with_ae / sum(subjects_with_ae), 1)
  )
```

::: {.callout-tip}
### Run It Yourself

Run the chunk and compare the last two columns.

The **correct** column divides by the safety population from ADSL — 200
subjects per arm — giving 75.5% on GLPX 10 mg and 55.0% on placebo.
Those are incidence rates: the proportion of *dosed subjects* who had an
event. A 20-point difference between arms, which is the finding.

The **wrong** column divides by the number of subjects who had events at
all, giving 57.9% and 42.1%. Those two figures sum to 100%, which is the
tell: it is not incidence at all but a *share of affected subjects*. It
answers "of the people who had events, what fraction were on drug?" —
a question nobody asked, that looks like the one they did ask, and that
mechanically shrinks the apparent gap between the arms.

The 138 subjects with no adverse events never appear in ADAE. They are
the entire difference, and they exist only in ADSL. This is why OCCDS
tells you to take denominators from ADSL, and why ADSL is required in a
submission even when nothing else is.
:::

### Conformance

Finally, check the dataset against what OCCDS says it should be:

```{r conformance}
#| eval: true

tibble(
  check = c("one row per source AE record",
            "AESEQ + USUBJID unique (datapoint traceability)",
            "every row joins to an ADSL subject",
            "TRTEMFL null only where ASTDT is null",
            "at most one AOCCFL per subject"),
  pass = c(
    nrow(adae) == nrow(ae),
    nrow(distinct(adae, USUBJID, AESEQ)) == nrow(adae),
    all(adae$USUBJID %in% adsl$USUBJID),
    all(is.na(adae$TRTEMFL) == is.na(adae$ASTDT)),
    max(tapply(adae$AOCCFL, adae$USUBJID,
               \(x) sum(x == "Y", na.rm = TRUE))) == 1
  )
)
```

All five pass. The first is the OCCDS structural claim — one record per
record in the source SDTM domain (OCCDS v1.1, §3.1.2) — and the second
is what makes each of those records traceable back to the exact SDTM row
it came from.

This dataset is written to `data/adam/adae.csv` by `R/build_adam.R`,
alongside ADSL, and the TLF sessions read it from there.

## What Can Go Wrong

*Estimated time: ~10 minutes (reading)*

**Computing percentages from the event dataset.** The single most
consequential error in this session, and the easiest to make because the
code runs and returns plausible numbers. Denominators come from ADSL
(OCCDS v1.1, §3.2.1). If a percentage's denominator was not counted in
ADSL, check it again.

**Treating `TRTEMFL` as an SDTM fact.** It is derived in ADaM, from
ADSL's dates and the sponsor's window. Where SDTM carries `AETRTEM`, the
two "may differ due to different definitions, date imputation, and other
analysis rules" (OCCDS v1.1, §3.2, Table 3.2.5.3) — and the analysis
uses the ADaM one.

**Coding an unclassifiable event as `"N"`.** A missing date means "we do
not know," not "not treatment emergent." Null is the honest value, and
it keeps the row visible as something requiring resolution.

**Imputing silently.** If you apply an imputation rule, `ASTDTF` must
record the level of imputation (OCCDS v1.1, §3.2, Table 3.2.4.1). An
imputed date with no flag is an assertion presented as an observation.

**Recoding `AESEV` in place.** Same-name variables must keep SDTM's
values (ADaMIG v1.3, §3.1.1, item 3). Recased or regrouped severity goes
in `ASEV`, `ASEVN` or `SEVGRy`.

**Expecting `PARAM` and `AVAL`.** They do not exist in OCCDS (OCCDS
v1.1, §3.2). Reaching for them is a sign the data might belong in BDS —
or that the analysis question has been misread.

**Claiming SubClass ADVERSE EVENT without the MedDRA hierarchy.** The
SubClass requires all five levels of the primary path (OCCDS v1.1,
§3.2.3). A dataset labelled with a SubClass it does not satisfy is a
metadata error a validator will find.

### Common misinterpretations

**"OCCDS is just BDS without the value column."** It is a different
structure for a different kind of question. BDS analyses measurements
against a baseline; OCCDS counts occurrences within a population. The
absence of `AVAL` follows from that, rather than defining it.

**"`AOCCFL` marks the first adverse event."** It marks the first
*treatment-emergent* one. Events outside the window are not candidates,
which is why the flags are derived after `TRTEMFL` and not before.

**"Uncoded data means we cannot build ADAE."** OCCDS says the opposite:
data that could have been coded but was not *should* use this structure
(OCCDS v1.1, §1.1.2). What is lost is the dictionary-keyed subset of
analyses, and CDISC tells you exactly which those are.

**"Every ADaM dataset is built from SDTM."** ADAE is built from SDTM AE
*and* from ADSL. Analysis datasets feed each other, and ADSL feeds
almost everything.

## Exercises

Practice: [Exercise 5 — ADAE](../exercises/adae.qmd).

## Comprehension Check

1. GLPX-001's ADAE has 389 rows covering 262 subjects, and ADSL has 400.
   Which number is the denominator for "percentage of subjects with a
   treatment-emergent adverse event," and why are the other two wrong?
2. `TRTEMFL` needs `ASTDT`, `TRTSDT` and `TRTEDT`. Which of those three
   come from ADSL, and what does that tell you about the order in which
   analysis datasets must be built?
3. Why is `ASTDTF` null for subject `103-199`'s undated event, when the
   date is certainly not known?
4. Your study's AE data is fully MedDRA coded. Which occurrence flags
   become available that GLPX-001 cannot derive, and where in OCCDS
   would you look to confirm the list?
5. A colleague recases `AESEV` from `MILD` to `Mild` directly, arguing
   the meaning is unchanged. Which rule does this break, and what should
   they do instead?

::: {.callout-note collapse="true"}
### Answers

1. **400**, the safety population from ADSL. 262 counts only subjects
   who had an event, which excludes the 138 event-free subjects from the
   denominator and inflates every percentage. 389 counts *events*, not
   subjects, and a subject with four events would be counted four times.
   OCCDS v1.1, §3.2.1 states the rule directly.
2. `TRTSDT` and `TRTEDT` come from ADSL; only `ASTDT` derives from the
   AE domain. SDTM AE has no knowledge of when the subject was dosed, so
   ADSL must exist before ADAE can be built. Analysis datasets are a
   dependency graph, and ADSL is at its root — which is why it is
   required in a submission even when nothing else is.
3. Because `ASTDTF` records the *level of imputation applied*, and
   nothing was imputed — this study specified no imputation rule
   (OCCDS v1.1, §3.2, Table 3.2.4.1). The flag describes what the
   derivation did, not what the data lacks. Populating it would claim an
   imputation that never happened.
4. `AOCCPFL` and `AOCCPIFL` (keyed on the preferred term, `--DECOD`),
   plus `AOCCSFL` and `AOCCSIFL` (keyed on system organ class). Look in
   **both** occurrence-flag tables, not just the one titled *MedDRA
   Occurrence Flag Variables*: `AOCCSFL` and `AOCCSIFL` are in Table
   3.2.6.2, but `AOCCPFL` and `AOCCPIFL` sit in the general Table
   3.2.6.1 and are dictionary-dependent all the same. The reliable test
   is what each derivation keys on, not which table it appears in.
   Coded data would also make SubClass ADVERSE EVENT reachable (OCCDS
   v1.1, §3.2.3).
5. "Same name, same meaning, same values" (ADaMIG v1.3, §3.1.1, item 3):
   a variable carrying an SDTM name must be a copy of the SDTM variable,
   values unmodified. It would also break the `AESEV` codelist (C66769),
   whose submission values are uppercase. The recased version belongs in
   `ASEV`, which is producer-defined terminology and carries no CDISC
   codelist to violate.
:::
